---
title: "SMDS Homework - Block 4"
author: "P. Morichetti, M. Rispoli, A. Cicchini and E. Ballarin  |  Group 'B'"
date: "27th May 2020"
output:
  html_document:
    theme: darkly
    highlight: breezedark
    mathjax: default
    self_contained: true
    md_extensions: +autolink_bare_uris
    toc: true
    toc_collapsed: false
    toc_float: true
    toc_depth: 3
    number_sections: false
header-includes:
- \usepackage{color}
- \usepackage{graphicx}
- \usepackage{grffile}
institute: University of Trieste, SISSA, ICTP, University of Udine
graphics: yes
fontsize: 10pt
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.align = 'center', warning=FALSE, message=FALSE, fig.asp=0.625, dev='png', global.par = TRUE, dev.args=list(pointsize=10), fig.path = 'figs/')
```

```{r setup, include=FALSE}
library(knitr)
local({
  hook_plot = knit_hooks$get('plot')
  knit_hooks$set(plot = function(x, options) {
    paste0('\n\n----\n\n', hook_plot(x, options))
  })
})
```


# Exercises from *DAAG, Chapter 6* 

## Exercise 6

### Text

### Solution

## Exercise 8

### Text

Apply the `lm.ridge()` function to the `litters` data, using the generalized cross-validation (GCV) criterion to choose the tuning parameter. (GCV is an approximation to cross-validation.)

a. In particular, estimate the coefficients of the model relating `brainwt` to `bodywt` and `lsize` and compare with the results obtained using `lm()`.  
b. Using both ridge and ordinary regression, estimate the mean brain weight when litter size is 10 and body weight is 7. Use the bootstrap, with case-resampling, to compute approximate 95% percentile confidence intervals using each method. Compare with the interval obtained using `predict.lm()`.

### Solution

```{r daag6_6b,code = readLines("src/daag6_8.R"), echo=TRUE}
```

## Exercise 10

### Text

### Solution


## Exercise 11

### Text

### Solution

# Exercises from *DAAG, Chapter 8* 

## Exercise 1

### Text

### Solution

## Exercise 2

### Text

### Solution


## Exercise 3

### Text

### Solution


## Exercise 6

### Text

### Solution